GADTs, Functoriality, Parametricity: Pick Two

### 2021-05-06 20-gadts/21-LSFA/GADT-story.tex
final LFSA paper

There are two standard ways to represent data types when studying
modern functional languages: {\em i}) as their Church encodings in a
(possibly higher-kinded) extension of System F~\cite{gir72}, the
calculus at the core of all such languages, and {\em ii}) by
augmenting System F with primitives for constructing them directly as
fixpoints.  Models in which the Church encoding and the fixpoint
representation of every algebraic data type (ADT) are semantically
equivalent include the operationally-based model of~\cite{pit98,pit00}
and the categorical models of~\cite{joh02,joh03,jgj21}.  In the
categorical model of~\cite{jgj21} this semantic equivalence also holds
for the syntactic generalization of ADTs known as nested
types~\cite{bm98}. Generalized algebraic data types (GADTs) generalize
nested types~\cite{pvww06} --- and thus further generalize ADTs ---
syntactically (see Section~\ref{sec:gadts} below).  Indeed, we have:

@@

\noindent
This sequence of syntactic generalizations suggests that a
corresponding sequence of semantic generalizations might also hold.
However, this is not the case. Specifically, the semantic equivalence
between Church encodings and fixpoint representations that holds for
ADTs and nested types need not hold for GADTs. In a language
supporting GADTs we must therefore choose whether to represent data
types as their Church encodings or as fixpoints.  The main result of
this paper shows that the representation chosen determines whether or
not a language supporting GADTs can have parametric
models~\cite{rey83}. It thus determines whether or not GADTs can enjoy
consequences of parametricity such as representation
independence~\cite{adr09,dnb12}, equivalences between
programs~\cite{hd11}, deep induction principles~\cite{jgj21s,jp20},
and useful (``free'') theorems about programs derived from their types
alone~\cite{wad89}.  This result is unintuitive, novel, and
surprising.

Consider the GADT $\mathsf{Seq}$ defined by\footnote{Although our
  development applies to GADTs in any language, we will use Haskell
  syntax for the code in this paper.}
\begin{equation}\label{eq:seq}
\begin{array}{l}
\mathsf{data\, Seq\,a\,where}\\
\mathsf{\;\;\;\;\;\;\;\;Const ::\, a \to Seq\,a}\\
\mathsf{\;\;\;\;\;\;\;\;Pair\,\,\,\,\, ::\, Seq \,a \to Seq\,b \to
  Seq\,(a \times b)}\\ 
\end{array}
\end{equation}
\noindent
This GADT comprises sequences of any type $\mathsf{a}$ and sequences
obtained by pairing the data in two already existing sequences. When
represented as its Church encoding, this GADT contains no data
elements other than these. More generally, a GADT's Church encoding
contains exactly those data elements that are representable by its
syntax. By contrast, the fixpoint representation of a GADT sees that
data type as the {\em functorial completion}~\cite{jp19} of its
syntax.  Completing a GADT's syntax from a funct{\em ion} on types to
a funct{\em or} on types is necessary if it is to be represented as a
fixpoint~\cite{tfca}, so functoriality is inherent in fixpoint
representations of data types. Since being a functor entails
supporting a map function satisfying the functor laws, the fixpoint
representation of a data type must include the entire ``map closure''
of its syntax.  For example, the fixpoint representation of
$\mathsf{Seq}$ contains not just those data elements representable by
its syntax, but also all data elements of the form
$\mathsf{map\,f\,s}$ for all types $\mathsf{t_1}$ and $\mathsf{t_2}$,
all functions $\mathsf{f : t_1 \to t_2}$ definable in the language and
all $\mathsf{s :: Seq\,t_1}$, as well as all data elements of the form
$\mathsf{map\,g\,s'}$ for each appropriately typed function
$\mathsf{g}$ and each element $\mathsf{s'}$ already added to the data
type, and so on. Functorial completion for $\mathsf{Seq}$ adds, in
particular, data elements of the form
$\mathsf{map\,g\,(Pair\,u_1\,u_2)}$ even though these may not
themselves be of the form $\mathsf{Pair\,v_1\,v_2}$ for any terms
$\mathsf{v_1}$ and $\mathsf{v_2}$. Importantly, functorial completion
adds no new data elements to the syntax of a GADT that is an ADT or
other nested type. This perfectly explains why Church encodings and
fixpoint representations coincide for ADTs and other nested
types. However, the two representations are not the same for GADTs
that are not nested types.\looseness=-1

@@

A GADT programmer is likely to use GADTs precisely {\em because} they
exhibit different behaviors at different types, and thus to consider a
GADT to be completely specified by its syntax. When used this way, the
shape of a particular element of a GADT is actually {\em determined}
by the data it contains. As a result, it cannot be expected to support
a functorial $\mathsf{map}$ function like ADTs and nested
types. Indeed, the definition for $\mathsf{Seq}$ above specifies that
an element of the form $\mathsf{Pair\,u_1\,u_2}$ must have the shape
of a sequence of data of pair type rather than a sequence of data of
arbitrary type $\mathsf{e}$.  The clause of $\mathsf{map}$ for the
$\mathsf{Pair}$ constructor should therefore feed $\mathsf{map}$ a
function $\mathsf{f :: (a \times b) \to e}$ and a term of the form
$\mathsf{Pair \,u_1\,u_2}$ for $\mathsf{u_1 :: Seq\,a}$ and
$\mathsf{u_2 :: Seq\,b}$, and produce a term $\mathsf{Pair\,v_1\,v_2}$
for some appropriately typed terms $\mathsf{v_1}$ and
$\mathsf{v_2}$. However, it is not clear how to achieve this since
$\mathsf{e}$ need not necessarily be a product type. And even if
$\mathsf{e}$ {\em were} known to be of the form $\mathsf{w \times z}$,
we still wouldn't necessarily have a way to produce data of type
$\mathsf{w \times z}$ from only $\mathsf{f :: a \times b \to w \times
  z}$ and $\mathsf{u_1}$ and $\mathsf{u_2}$ unless we knew, e.g., that
$\mathsf{f}$ was a pair of functions $\mathsf{(f_1 :: a \to w, f_2 ::
  b \to z)}$. When it is intended to capture this kind of
non-uniformity, a GADT cannot be regarded as a data-independent
container that can be filled with data of any type in the same way
that ADTs and nested types can. In this situation, a GADT cannot be
seen as a functor, and so must necessarily be represented as its
Church encoding.

A semanticist, on the other hand, is likely to expect GADTs to
generalize ADTs semantically --- i.e., to have the same kind of
functorial semantics that ADTs and nested types
have~\cite{bm98,jg07,jp19} --- and thus to be useable as a container
filled with data that can be changed without changing the shape of the
container itself.  For a GADT to be used in this way, its syntax must
reflect functoriality. Since the Church encoding of a GADT that is not
a nested type is not a functor, such a GADT must be viewed as its
fixpoint representation. The functorial completion inherent in the
fixpoint representation of a GADT adds to its syntax those, and only
those, data elements needed to support a (functorial) $\mathsf{map}$
function.  For example, let $\mathsf{1}$ be the unit type whose single
element is also denoted by $\mathsf{1}$, and let $\mathsf{G}$ be the
GADT defined by\looseness=-1
\begin{equation}\label{eq:G}
\begin{array}{l}
\mathsf{data\,G\,a\,where}\\
\mathsf{\;\;\;\;\;\;\;\;C :: G\,1}
\end{array}
\end{equation}
Then the functorial completion of $\mathsf{G}$ includes elements at
any instance $\mathsf{G\,a}$ for any type $\mathsf{a}$ that is not the
empty type $\mathsf{0}$ (since there is always a function from
$\mathsf{1}$ to such an $\mathsf{a}$), but includes no elements at
instance $\mathsf{G\,0}$. Indeed, $\mathsf{G\,0}$ is not inhabited via
the syntactic specification of $\mathsf{G}$, and is not inhabited via
functorial completion because it is not possible to define a function
from $\mathsf{1}$ into $\mathsf{0}$. Note, however, that since not all
instances of the fixpoint representation of a GADT need be inhabited,
fixpoint representations of GADTs are not, in general, nested types.

The key observation of this paper is that, while the viewpoints of the
GADT programmer and the semanticist are both valid, the two are
irreconcilable. Importantly, which representation of GADTs is adopted
in any particular setting has significant consequences for the ways
they can be used and reasoned about there. In particular, the way that
GADTs are represented has deep implications for parametric reasoning
about them. Specifically, a programmer who views GADTs as their Church
encodings cannot safely use program transformations or reasoning
principles that involve $\mathsf{map}$ functions for them, although
they may be able to program with and reason about GADTs using other
consequences of parametricity, such as type inhabitation results. On
the other hand, a semanticist who views GADTs as fixpoints will have
all naturality-based program transformations and reasoning principles
for GADTs at their disposal since these all derive from
functoriality. But since, as we show in Section~\ref{sec:par}, no
parametric model can be constructed for fixpoint representations of
GADTs, non-naturality consequences of parametricity will not
necessarily hold for them.  Overall, we show that, as with software
engineering's iron triangle~\cite{iron-triangle}, we can have any two
of {\em GADTs}, {\em functoriality}, and {\em parametricity} we like,
but we cannot have all three.

The goal of this paper is to show how the above observations can be
made precise. We thereby answer the question we posed above in the
affirmative: Yes, it really does matter how we represent GADTs.

@@

Since the shape of an ADT structure --- i.e., a structure whose type
is an instance of an ADT --- is independent of the type of data it
contains, ADTs can be defined polymorphically. As a result, an ADT
structure containing data of type $\mathsf{a}$ can be transformed into
another ADT structure of the exact same shape containing data of
another type $\mathsf{b}$ simply by applying a given function
$\mathsf{f \,:\,a \to b}$ to each of its elements. Indeed, every ADT
$\mathsf{T}$ can be made an instance of Haskell's $\mathsf{Functor}$
class by defining a type-and-data-uniform, structure-preserving,
data-changing function $\mathsf{map_T}$ for it.\footnote{We write
  $\mathsf{map_T}$, or simply $\mathsf{map}$ when $\mathsf{T}$ is
  clear from context, for the function $\mathsf{fmap :: (a \to b) \to
    (T\,a \to T\,b)}$ witnessing that a type constructor $\mathsf{T}$
  is an instance of Haskell's $\mathsf{Functor}$ class. This
  emphasizes that $\mathsf{fmap}$ functions in Haskell are intended to
  satisfy syntactic reflections of the functor laws ---
%  iality properties ---
  i.e., preservation of identity functions and composition of
  functions --- even though this is not enforced by the compiler and
  is instead left to the good intentions of the programmer.}
Moreover, given a type-independent way of rearranging an ADT
structure's shape $\mathsf{T\,a}$ into the shape for another ADT
structure $\mathsf{T'\,a}$, we get the same structure of type
$\mathsf{T'\,b}$ regardless of whether we first rearrange the original
structure of type $\mathsf{T\,a}$ into one of type $\mathsf{T'\,a}$
and then use $\mathsf{map_{T'}}$ to convert that resulting structure
to one of type $\mathsf{T'\,b}$, or we first use $\mathsf{map_T}$ to
convert the original structure of type $\mathsf{T\,a}$ to one of type
$\mathsf{T\,b}$ and then rearrange that resulting structure into one
of type $\mathsf{T'\,b}$. For example, if $\mathsf{f : a \to b}$,
$\mathsf{t \,:\,Tree\,a}$, and $\mathsf{g : Tree\,a\to List\,a}$
arranges lists into trees in a type-independent way, then we have the
following rearrange-transform property:
\[\mathsf{map_{List}\,f\,(g \, t) \,=\, g\,(map_{Tree}\,f\,t)}\]

@@

Church encodings of ADTs can be defined in any language that supports
functions. They can therefore be used to represent ADTs in languages
that do not support primitives for sum types, product types, or
recursion. The Church encodings of the ADTs $\mathsf{List\,a}$ and
$\mathsf{Tree\,a}$, for example, are
\[\mathsf{List\,a} = \mathsf{\forall b.\,b \to (a \to b \to b) \to b}\]
and 
\[\mathsf{Tree\,a} = \mathsf{\forall b.\,(a \to b) \to (b \to a \to b
  \to b) \to b}\] respectively. The argument types are abstractions of
the types of ADTs' constructors. For instance, $\mathsf{b}$ abstracts
the type $\mathsf{List\,a}$ of the constructor $\mathsf{Nil}$ for
lists, and $\mathsf{a \to b \to b}$ abstracts the type $\mathsf{a \to
  List\,a \to List\,a}$ of the constructor $\mathsf{Cons}$.

Because the types of the ``abstract constructors'' for an ADT are
uniform in their argument types, it is always possible to Church
encode the type constructors themselves as well. For example, the
Church encodings of the type constructors $\mathsf{List}$ and
$\mathsf{Tree}$ are
\[\mathsf{List} = \mathsf{\forall a. \forall b.\,b \to (a \to b \to b)
  \to b}\] 
and 
\[\mathsf{Tree} = \mathsf{\forall a. \forall b.\,(a \to b) \to (b \to
  a \to b \to b) \to b}\] respectively. This is what allows an ADT's
associated type constructor to be made an instance of Haskell's
$\mathsf{Functor}$ class. For example, the
$\mathsf{map}$ function
from Haskell's standard library
makes the type constructor $\mathsf{List}$ an
instance of the $\mathsf{Functor}$ class, and for $\mathsf{Tree}$ we
can define
\[\begin{array}{lll}
\mathsf{map_{Tree}} & \mathsf{::} & \mathsf{(a \to b) \to Tree\,a \to Tree \,b}\\
\mathsf{map_{Tree}\, f\, (Leaf\,x)} & \mathsf{=} & \mathsf{Leaf\,(f\,x)}\\
\mathsf{map_{Tree}\, f\, (Node\,t_1\,x\,t_2)} & \mathsf{=} &
\mathsf{Node\,(map_{Tree}\, f \,t_1)\,(f\,x)\,(map_{Tree}\,f\,t_2)}\\
\end{array}\]
Note, however, that the Church encoding of an ADT carries with it no
expectation whatsoever that such a type-and-data-uniform,
structure-preserving, data-changing $\mathsf{map}$ function can be
defined or that, if one can, it will satisfy a rearrange-transform
property. But such a $\mathsf{map}$ function can always be defined for
any ADT precisely because its Church encoding and its fixpoint
representation necessarily coincide, and this $\mathsf{map}$ function
will, indeed, satisfy a rearrange-transform property.

@@

Fixpoint representations capture in syntax the fact that ADTs can be
considered as fixpoints of functors. For example $\mathsf{List\,a}$ is
indeed a fixpoint of $\mathsf{F_{List\,a}}$, where
$\mathsf{F_{List\,a}\,X} = \mathsf{1 + a \times X}$ as determined
by~\eqref{eq:list}, i.e., is indeed such that
\[\mathsf{List\,a} = \mathsf{1 + a \times List\,a}\] 
since every element of $\mathsf{List\,a}$ is either empty or is
obtained by consing an element of type $\mathsf{a}$ onto an
already-existing structure of type $\mathsf{List\,a}$. In fact, this
fixpoint equation is just a rewriting of the Haskell data type
declaration for $\mathsf{List\,a}$. We therefore have that
$\mathsf{List\, a} = \mathsf{\mu F_{List\,a}}$ is modeled by $\mu
F_{\mathit{List}\,a}$, where the functor $F_{\mathit{List}\,a}$ models
the type constructor $\mathsf{F_{List\,a}}$.\footnote{Throughout, we
  use $\mathtt{sans\; serif}$ font for program text and $\mathit{math
    \; italic}$ font for semantic objects.}  Similarly, the ADT
$\mathsf{Tree\,a}$ can be seen to be a fixpoint of
$\mathsf{F_{Tree\,a}}$, where $\mathsf{F_{Tree\,a}\,X} = \mathsf{a + X
  \times a \times X}$ by~\eqref{eq:tree}, so that $\mathsf{Tree\, a} =
\mathsf{ \mu F_{Tree\,a}}$ is modeled by $\mu F_{\mathit{Tree}\,a}$,
where $F_{\mathit{Tree}\,a} $ models $\mathsf{F_{Tree\,a}}$.

@@

Requiring $F$ to be a functor ensures that the interpretation $\mu F$
of the ADT $\mathsf{T\,a} = \mathsf{\mu F}$ exists. But to ensure that
$\mu F$ is itself a functor, so that the type constructor $\mathsf{T}$
associated with $\mathsf{T\,a}$ also supports its own $\mathsf{map}$
function, we can require that $F$ also be the fixpoint $\mu H$ of a
functor $H$. In this case, however, $H$ must be a functor on the
category $\mathit{Set}^\mathit{Set}$ of functors and natural
transformations on $\mathit{Set}$. That is, $H$ must be a {\em
  higher-order} functor on $\mathit{Set}$. Reflecting this requirement
back into syntax gives that $\mathsf{T\,a} = \mathsf{(\mu H)\,a}$ for
the ``type constructor constructor'' $\mathsf{H}$ that supports
suitable\footnote{The $\mathsf{map}$ function for $\mathsf{H}$ is
  intended to satisfy syntactic reflections of the functor laws in
  $\mathit{Set}^\mathit{Set}$ --- i.e., preservation of identity
  natural transformations and composition of natural transformations
  --- and the $\mathsf{map}$ function for $\mathsf{H\,F}$ is intended
  to satisfy syntactic reflections of the functor laws in
  $\mathit{Set}$ even though there is no mechanism in Haskell for
  enforcing this.} $\mathsf{map}$ functions.

@@

In any parametric model, the interpretations of the Church encoding
and the fixpoint representation of an ADT or a nested type will
necessarily be isomorphic. But whereas it is impossible even to {\em
  state} the rearrange-transform property for their Church encodings
(unless $\mathsf{map}$ functions have somehow been defined for them),
for their fixpoint representations such a property is simply the
reflection back into syntax of the instance of naturality for the
type-independent function that rearranges structures of type
$\mathsf{T\,a}$ into ones of type $\mathsf{T'\,a}$ and the
content-transforming functions $\mathsf{map_T\,f}$ and
$\mathsf{map_{T'}\,f}$ for a function $\mathsf{f :: a \to b}$, where
$\mathsf{T}$ and $\mathsf{T'}$ are the type constructors associated
with these ADTs, respectively.

@@

In addition to not being semantically equivalent, the two
representations of GADTs have very different implications for
parametricity. We explore the differences in parametricity results for
GADTs with respect to the two representations in the next section.

>>>>

There are two standard ways to represent data types when studying
modern functional languages: {\em i}) as their Church encodings in a
(possibly higher-kinded) extension of System F~\cite{gir72}, the
calculus at the core of all such languages, and {\em ii}) by
augmenting System F with primitives for constructing them directly as
fixpoints.  Models in which the Church encoding and the fixpoint
representation of every algebraic data type (ADT) are semantically
equivalent include the operationally-based model of~\cite{pit98,pit00}
and the categorical models of~\cite{joh02,joh03,jgj21}.  In the
categorical model of~\cite{jgj21} this semantic equivalence also holds
for the syntactic generalization of ADTs known as nested
types~\cite{bm98}. Generalized algebraic data types
(GADTs)~\cite{pvww06} generalize nested types --- and thus further
generalize ADTs --- syntactically (see Section~\ref{sec:gadts} below):

@@

\noindent
This sequence of syntactic generalizations suggests that a
corresponding sequence of semantic generalizations might also hold.
However, this is not the case. Specifically, the semantic equivalence
between Church encodings and fixpoint representations that holds for
ADTs and nested types need not hold for GADTs. In a language
supporting GADTs we must therefore choose whether to represent data
types as their Church encodings or as fixpoints.  The main result of
this paper shows that the choice of representation of data types in a
language supporting GADTs determines whether or not that language can
have parametric models~\cite{rey83}. It thus determines whether or not
GADTs can enjoy consequences of parametricity such as representation
independence~\cite{adr09,dnb12}, equivalences between
programs~\cite{hd11}, deep induction principles~\cite{jgj21s,jp20},
and useful (``free'') theorems about programs derived from their types
alone~\cite{wad89}.  This result is unintuitive, novel, and
surprising.

Consider the GADT $\mathsf{Seq}$ defined by\footnote{Although our
  development applies to GADTs in any language, we will use Haskell
  syntax for the code in this paper.}
\begin{equation}\label{eq:seq}
\begin{array}{l}
\mathsf{data\, Seq\,a\,where}\\
\mathsf{\;\;\;\;\;\;\;\;Const ::\, a \to Seq\,a}\\
\mathsf{\;\;\;\;\;\;\;\;Pair\,\,\,\,\, ::\, Seq \,a \to Seq\,b \to
  Seq\,(a \times b)}\\ 
\end{array}
\end{equation}
\noindent
This GADT comprises sequences of any type $\mathsf{a}$ and sequences
obtained by pairing the data in two already existing sequences. When
represented as its Church encoding, this GADT contains no data
elements other than these. More generally, a GADT's Church encoding
contains exactly those data elements that are representable by its
syntax. By contrast, the fixpoint representation of a GADT sees that
data type as the {\em functorial completion}~\cite{jp19} of its
syntax.  Completing a GADT's syntax from a funct{\em ion} on types to
a funct{\em or} on types is necessary if its interpretation is to be a
fixpoint~\cite{tfca}, so functoriality is inherent in fixpoint
representations of data types. Since being a functor entails
supporting a map function satisfying the functor laws, the fixpoint
representation of a data type must include the entire ``map closure''
of its syntax.  For example, the fixpoint representation of
$\mathsf{Seq}$ contains not just those data elements representable by
its syntax, but also all data elements of the form
$\mathsf{map\,f\,s}$ for all types $\mathsf{t_1}$ and $\mathsf{t_2}$,
all functions $\mathsf{f :: t_1 \to t_2}$ definable in the language and
all $\mathsf{s :: Seq\,t_1}$, as well as all data elements of the form
$\mathsf{map\,g\,s'}$ for each appropriately typed function
$\mathsf{g}$ and each element $\mathsf{s'}$ already added to the data
type, and so on. Functorial completion for $\mathsf{Seq}$ adds, in
particular, data elements of the form
$\mathsf{map\,g\,(Pair\,u_1\,u_2)}$ even though these may not
themselves be of the form $\mathsf{Pair\,v_1\,v_2}$ for any terms
$\mathsf{v_1}$ and $\mathsf{v_2}$. Importantly, functorial completion
adds no new data elements to the syntax of a GADT that is an ADT or
other nested type, which perfectly explains why Church encodings and
fixpoint representations coincide for ADTs and other nested
types. However, the two representations are not the same for GADTs
that are not nested types.\looseness=-1

@@

A GADT programmer is likely to use GADTs precisely {\em because} they
exhibit different behaviors at different types, and thus to consider a
GADT to be completely specified by its syntax. When used this way, the
shape of a particular element of a GADT is actually {\em determined}
by the data it contains. As a result, it cannot be expected to support
a functorial $\mathsf{map}$ function like ADTs and nested
types. Indeed, the definition for $\mathsf{Seq}$ above specifies that
an element of the form $\mathsf{Pair\,u_1\,u_2}$ must have the shape
of a sequence of data of pair type rather than a sequence of data of
arbitrary type $\mathsf{e}$.  The clause of $\mathsf{map}$ for the
$\mathsf{Pair}$ constructor should therefore feed $\mathsf{map}$ a
function $\mathsf{f :: (a \times b) \to e}$ and a term of the form
$\mathsf{Pair \,u_1\,u_2}$ for $\mathsf{u_1 :: Seq\,a}$ and
$\mathsf{u_2 :: Seq\,b}$, and produce a term $\mathsf{Pair\,v_1\,v_2}$
for some appropriately typed terms $\mathsf{v_1}$ and
$\mathsf{v_2}$. However, it is not clear how to achieve this since
$\mathsf{e}$ need not necessarily be a product type. And even if
$\mathsf{e}$ {\em were} known to be of the form $\mathsf{w \times z}$,
we still wouldn't necessarily have a way to produce data of type
$\mathsf{w \times z}$ from only $\mathsf{f :: a \times b \to w \times
  z}$ and $\mathsf{u_1}$ and $\mathsf{u_2}$ unless we knew, e.g., that
$\mathsf{f}$ was a pair of functions $\mathsf{(f_1 :: a \to w, f_2 ::
  b \to z)}$. When it is intended to capture this kind of
non-uniformity, a GADT cannot be regarded as a data-independent
container that can be filled with data of any type in the same way
that ADTs and nested types can. In this situation, a GADT cannot
denote a functor, and so must necessarily be represented as its
Church encoding.

A semanticist, on the other hand, is likely to expect GADTs to
generalize ADTs semantically --- i.e., to have the same kind of
functorial semantics that ADTs and nested types
have~\cite{bm98,jg07,jp19} --- and thus to be useable as a container
filled with data that can be changed without changing the shape of the
container itself.  For a GADT to be used in this way, its syntax must
reflect functoriality. Since the Church encoding of a GADT that is not
a nested type does not denote a functor, to have a functorial
semantics such a GADT must be viewed as its fixpoint
representation. The functorial completion inherent in the fixpoint
representation of a GADT adds to its syntax those, and only those,
data elements needed to support a functorial $\mathsf{map}$ function.
For example, let $\mathsf{1}$ be the unit type whose single element is
also denoted by $\mathsf{1}$, and let $\mathsf{G}$ be the GADT defined
by\looseness=-1
\begin{equation}\label{eq:G}
\begin{array}{l}
\mathsf{data\,G\,a\,where}\\
\mathsf{\;\;\;\;\;\;\;\;C :: G\,1}
\end{array}
\end{equation}
Then the functorial completion of $\mathsf{G}$ includes elements at
any instance $\mathsf{G\,a}$ for any type $\mathsf{a}$ that is not the
empty type $\mathsf{0}$ (since there is always a function from
$\mathsf{1}$ to such an $\mathsf{a}$), but includes no elements at
instance $\mathsf{G\,0}$. Indeed, $\mathsf{G\,0}$ is not inhabited via
the syntactic specification of $\mathsf{G}$, and is not inhabited via
functorial completion because it is not possible to define a function
from $\mathsf{1}$ into $\mathsf{0}$.
%Note, however, that since not all
%instances of the fixpoint representation of a GADT need be inhabited,
%fixpoint representations of GADTs are not, in general, nested types.

The key observation of this paper is that, while the viewpoints of the
GADT programmer and the semanticist are both valid, the two are
irreconcilable. Importantly, which of our two representations of data
types is adopted in any particular setting has significant
consequences for the ways GADTs can be used and reasoned about
there. In particular, the way that GADTs are represented has deep
implications for parametric reasoning about them. Specifically, a
programmer who views GADTs as their Church encodings cannot safely use
program transformations or reasoning principles that involve
$\mathsf{map}$ functions for them, although they may be able to
program with and reason about GADTs using other consequences of
parametricity, such as type inhabitation results. On the other hand, a
semanticist who views GADTs as fixpoints will have all
naturality-based program transformations and reasoning principles for
GADTs at their disposal since these all derive from functoriality. But
since, as we show in Section~\ref{sec:par}, no parametric model can be
constructed for fixpoint representations of GADTs, non-naturality
consequences of parametricity will not necessarily hold for them.
Overall, we show that, as with software engineering's iron
triangle~\cite{iron-triangle}, we can have any two of {\em GADTs},
{\em functoriality}, and {\em parametricity} we like, but we cannot
have all three.

The goal of this paper is to show how the above observations can be
made precise, and thereby to answer the question we posed above in the
affirmative: Yes, it really does matter how we represent GADTs.

@@

Since the shape of an ADT structure --- i.e., a structure whose type
is an instance of an ADT --- is independent of the type of data it
contains, ADTs can be defined polymorphically. As a result, an ADT
structure containing data of type $\mathsf{a}$ can be transformed into
another ADT structure of the exact same shape containing data of
another type $\mathsf{b}$ simply by applying a given function
$\mathsf{f \,::\,a \to b}$ to each of its elements. Moreover, every ADT
$\mathsf{T}$ can be made an instance of Haskell's $\mathsf{Functor}$
class by defining a type-and-data-uniform, structure-preserving,
data-changing function $\mathsf{map_T}$ for it.\footnote{We write
  $\mathsf{map_T}$, or simply $\mathsf{map}$ when $\mathsf{T}$ is
  clear from context, for the function $\mathsf{fmap :: (a \to b) \to
    (T\,a \to T\,b)}$ witnessing that a type constructor $\mathsf{T}$
  is an instance of Haskell's $\mathsf{Functor}$ class. We
  emphasize that $\mathsf{fmap}$ functions in Haskell are intended to
  satisfy syntactic reflections of the functor laws ---
%  iality properties ---
  i.e., preservation of identity functions and composition of
  functions --- even though this is not enforced by the compiler and
  is instead left to the good intentions of the programmer.}
Then, given a type-independent way of rearranging an ADT
structure's shape $\mathsf{T\,a}$ into the shape for another ADT
structure $\mathsf{T'\,a}$, we get the same structure of type
$\mathsf{T'\,b}$ regardless of whether we first rearrange the original
structure of type $\mathsf{T\,a}$ into one of type $\mathsf{T'\,a}$
and then use $\mathsf{map_{T'}}$ to convert that resulting structure
to one of type $\mathsf{T'\,b}$, or we first use $\mathsf{map_T}$ to
convert the original structure of type $\mathsf{T\,a}$ to one of type
$\mathsf{T\,b}$ and then rearrange that resulting structure into one
of type $\mathsf{T'\,b}$. For example, if $\mathsf{f :: a \to b}$,
$\mathsf{t \,::\,Tree\,a}$, and $\mathsf{g :: Tree\,a\to List\,a}$
arranges lists into trees in a type-independent way, then we have the
following rearrange-transform property:
\[\mathsf{map_{List}\,f\,(g \, t) \,=\, g\,(map_{Tree}\,f\,t)}\]

@@

Church encodings of ADTs can be defined in any language that supports
functions. They can therefore be used to represent ADTs in languages
that do not support primitives for sum types, product types, or
recursion. The Church encodings of the ADTs $\mathsf{List\,a}$ and
$\mathsf{Tree\,a}$, for example, are
\[\mathsf{List\,a} = \mathsf{\forall b.\,b \to (a \to b \to b) \to b}\]
and 
\[\mathsf{Tree\,a} = \mathsf{\forall b.\,(a \to b) \to (b \to a \to b
  \to b) \to b}\] respectively. The argument types in a Church
encoding of an ADT are abstractions of the types of the ADT's
constructors. For instance, $\mathsf{b}$ abstracts the type
$\mathsf{List\,a}$ of the constructor $\mathsf{Nil}$ for lists, and
$\mathsf{a \to b \to b}$ abstracts the type $\mathsf{a \to List\,a \to
  List\,a}$ of the constructor $\mathsf{Cons}$.

Because the types of the ``abstract constructors'' for an ADT are
uniform in their argument types, it is always possible to Church
encode the type constructors themselves as well. For example, the
Church encodings of the type constructors $\mathsf{List}$ and
$\mathsf{Tree}$ are
\[\mathsf{List} = \mathsf{\forall a. \forall b.\,b \to (a \to b \to b)
  \to b}\] 
and 
\[\mathsf{Tree} = \mathsf{\forall a. \forall b.\,(a \to b) \to (b \to
  a \to b \to b) \to b}\] respectively. This fact is what allows an
ADT's associated type constructor to be made an instance of Haskell's
$\mathsf{Functor}$ class. For example, the $\mathsf{map}$ function
from Haskell's standard library makes the type constructor
$\mathsf{List}$ an instance of the $\mathsf{Functor}$ class, and for
$\mathsf{Tree}$ we can define
\[\begin{array}{lll}
\mathsf{map_{Tree}} & \mathsf{::} & \mathsf{(a \to b) \to Tree\,a \to Tree \,b}\\
\mathsf{map_{Tree}\, f\, (Leaf\,x)} & \mathsf{=} & \mathsf{Leaf\,(f\,x)}\\
\mathsf{map_{Tree}\, f\, (Node\,t_1\,x\,t_2)} & \mathsf{=} &
\mathsf{Node\,(map_{Tree}\, f \,t_1)\,(f\,x)\,(map_{Tree}\,f\,t_2)}\\
\end{array}\]
Note that the Church encoding of an ADT carries with it no expectation
whatsoever that such a type-and-data-uniform, structure-preserving,
data-changing $\mathsf{map}$ function can be defined or that, if one
can, it will satisfy a rearrange-transform property. (Of course, such
a $\mathsf{map}$ function can always be defined for any ADT precisely
because its Church encoding and its fixpoint representation coincide,
and this $\mathsf{map}$ function will necessarily satisfy a
rearrange-transform property.)

@@

Fixpoint representations capture in syntax the fact that ADTs can be
considered as fixpoints of functors. For example, $\mathsf{List\,a}$ is
indeed a fixpoint of $\mathsf{F_{List\,a}}$, where
$\mathsf{F_{List\,a}\,X} = \mathsf{1 + a \times X}$
by~\eqref{eq:list}. That is,
\[\mathsf{List\,a} = \mathsf{1 + a \times List\,a}\] 
since every element of $\mathsf{List\,a}$ is either empty or is
obtained by consing an element of type $\mathsf{a}$ onto an
already-existing structure of type $\mathsf{List\,a}$. In fact, this
fixpoint equation is just a rewriting of the Haskell data type
declaration for $\mathsf{List\,a}$. We therefore have that
$\mathsf{List\, a} = \mathsf{\mu F_{List\,a}}$ is modeled by $\mu
F_{\mathit{List}\,a}$, where the functor $F_{\mathit{List}\,a}$ models
the type constructor $\mathsf{F_{List\,a}}$.\footnote{Throughout this
  paper, we use $\mathtt{sans\; serif}$ font for program text and
  $\mathit{math \; italic}$ font for semantic objects.}  Similarly,
the ADT $\mathsf{Tree\,a}$ can be seen to be a fixpoint of
$\mathsf{F_{Tree\,a}}$, where $\mathsf{F_{Tree\,a}\,X} = \mathsf{a + X
  \times a \times X}$ by~\eqref{eq:tree}, so that $\mathsf{Tree\, a} =
\mathsf{ \mu F_{Tree\,a}}$ is modeled by $\mu F_{\mathit{Tree}\,a}$ if
$F_{\mathit{Tree}\,a} $ models $\mathsf{F_{Tree\,a}}$.

@@

Requiring $F$ to be a functor ensures that the interpretation $\mu F$
of the ADT $\mathsf{T\,a} = \mathsf{\mu F}$ exists. But to ensure that
$\mu F$ is itself a functor, so that the type constructor $\mathsf{T}$
associated with $\mathsf{T\,a}$ also supports its own $\mathsf{map}$
function, we can require that $F$ also be the fixpoint $\mu H$ of a
functor $H$. In this case, however, $H$ must be a functor on the
category $\mathit{Set}^\mathit{Set}$ of functors and natural
transformations on $\mathit{Set}$. That is, $H$ must be a {\em
  higher-order} functor on $\mathit{Set}$. Reflecting this requirement
back into syntax gives that $\mathsf{T\,a} = \mathsf{(\mu H)\,a}$ for
the ``type constructor constructor'' $\mathsf{H}$ that supports
suitable\footnote{The $\mathsf{map}$ function for $\mathsf{H}$ is
  intended to satisfy syntactic reflections of the functor laws in
  $\mathit{Set}^\mathit{Set}$ --- i.e., preservation of identity
  natural transformations and composition of natural transformations
  --- and the $\mathsf{map}$ function for $\mathsf{H\,F}$ is intended
  to satisfy syntactic reflections of the functor laws in
  $\mathit{Set}$, even though there is no mechanism in Haskell for
  enforcing this.} $\mathsf{map}$ functions.

@@

In any parametric model, the Church encoding and the fixpoint
representation of an ADT or a nested type will necessarily be
semantically equivalent. But whereas it is impossible even to {\em
  state} the rearrange-transform property for their Church encodings
(unless functorial $\mathsf{map}$ functions have somehow been defined
for them), for their fixpoint representations such a property is
simply the reflection back into syntax of the instance of naturality
for the type-independent function that rearranges structures of type
$\mathsf{T\,a}$ into ones of type $\mathsf{T'\,a}$ and the
structure-preserving, data-changing functions $\mathsf{map_T\,f}$ and
$\mathsf{map_{T'}\,f}$ for a function $\mathsf{f :: a \to b}$, where
$\mathsf{T}$ and $\mathsf{T'}$ are the type constructors associated
with these ADTs, respectively.

@@

Because they are not semantically equivalent, the two representations
of GADTs have very different implications for parametricity. We
explore the differences in parametricity results for the two
representations of GADTs in the next section.
